---
title: "Class 'dm' and basic operations"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Class 'dm' and basic operations}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
fansi::set_knit_hooks(knitr::knit_hooks)
options(crayon.enabled = TRUE, width = 75, cli.width = 75)

knit_print.grViz <- function(x, ...) {
  x %>% 
    DiagrammeRsvg::export_svg() %>% 
    c("`````{=html}\n", ., "\n`````\n") %>% 
    knitr::asis_output()
}
```

The goal of the {dm} package and the `dm` class that comes with it, is to make your life easier when you are dealing with data and data models.

Let's take a look at the `dm` class.

## Class `dm`

The `dm` class contains a set of tables, as well as information about their primary keys and the relations between the tables. 
You can have your `dm` object based on different data sources, such as a local environment or a database.

`dm` is a wrapper around two classes that are independent from each other:

1. one part of the object is a [`src` object](https://dplyr.tidyverse.org/reference/src.html) from [{dplyr}](https://dplyr.tidyverse.org/)
2. the other part is an object of the `data_model` class
 from the [{datamodelr}](https://github.com/bergant/datamodelr) package

The `src` object contains information about where the tables of your data model are physically stored. 
The `data_model` object contains meta-information about your data, such as:

- what the primary keys in your schema are
- how the tables are related to each other, i.e. what the foreign key relations are

Last but not least, the third part of the object is a named list that contains the tables on the `src`.

So much about theory, let us now see how to construct such an object and how it looks like in `R`:

## Examples of `dm` objects {#ex_dm}

The relevant functions for creating `dm` objects are:

1. `dm()`
2. `as_dm()`
3. `new_dm()`

We will use tables from the popular  [{nycflights13}](https://github.com/hadley/nycflights13) package:
```{r message=FALSE}
library(dm)
library(dplyr)
flights_dm <- dm(src_df(pkg = "nycflights13"))
flights_dm
```

Here we are making use of the fact that the function `dm(src, data_model = NULL)` includes all available tables on a source in the `dm` object, should the `data_model`-argument be left `NULL`. 
This means that you can use this, for example, on a postgres database that you access via `src_postgres()` (with the appropriate arguments `dbname`, `host`, `port`, ...), to produce a `dm` object with all the tables on the database.

Another way of creating a `dm` object is calling `as_dm()` on a list of `tbls`, e.g., the tables included in the well-known [`iris`](https://www.rdocumentation.org/packages/datasets/versions/3.6.1/topics/iris) dataset:
```{r}
iris_dm <- as_dm(list("iris1" = iris, "iris2" = iris))
iris_dm
```

<!-- And lastly, with `new_dm()` you can create a `dm` by providing the three individual parts that it consists of: -->
<!-- ```{r} -->
<!-- local_src <- src_df(env = new.env()) -->
<!-- tables <- list("iris1" = iris, "iris2" = iris) -->
<!-- data_model <- cdm_get_data_model(iris_dm) -->
<!-- another_iris_dm <-  -->
<!--   new_dm( -->
<!--     tables = tables, -->
<!--     data_model = data_model -->
<!--     ) -->
<!-- another_iris_dm -->
<!-- ``` -->

<!--You noticed above that by using the `cdm_get_data_model()` function, we were able to access the `data_model` part of a `dm`.-->

We can get the list of tables with `cdm_get_tables()` and the `src` object with `cdm_get_src()`.

In order to pull a specific table from a `dm`, use:
```{r}
tbl(flights_dm, "airports")
``` 

But how can we use {dm}-functions to manage the primary keys of the tables in a `dm` object?

## Primary keys of `dm` objects {#pk}

Some useful functions for managing primary key settings are:

1. `cdm_add_pk()`
2. `cdm_has_pk()`
3. `cdm_get_pk()`
4. `cdm_rm_pk()`
5. `cdm_enum_pk_candidates()`
6. `cdm_get_all_pks()`

Currently `dm` objects only support one-column primary keys. If your tables have unique compound keys, adding a surrogate key column might be helpful. 
If you created a `dm` object according to the examples in ["Examples of `dm` objects"](#ex_dm), your object does not yet have any primary keys set.
So let's add one.

`cdm_add_pk()` has an option to check if the column of the table given by the user is a unique key (for performance reasons check is turned off by default).
Since the `iris` dataset does not have any unique one-column keys, we will use the `nycflights13` tables, i.e. `flights_dm` from above.

```{r}
cdm_has_pk(flights_dm, airports)
flights_dm_with_key <- cdm_add_pk(flights_dm, airports, faa)
flights_dm_with_key
```
The `dm` model now has a primary key. Let's check:
```{r}
cdm_has_pk(flights_dm_with_key, airports)
```
Get the name of the column that is marked as primary key of the table:
```{r}
cdm_get_pk(flights_dm_with_key, airports)
```
Remove a primary key:
```{r}
cdm_rm_pk(flights_dm_with_key, airports) %>% 
  cdm_has_pk(airports)
```

If you still need to get to know your data better, and it is already available in the form of a `dm` object, you can use the  `cdm_enum_pk_candidates()` function in order to get information about which columns of the table are unique keys:
```{r}
cdm_enum_pk_candidates(flights_dm_with_key, airports)
```
The `flights` table does not have any one-column primary key candidates:
```{r}
cdm_enum_pk_candidates(flights_dm_with_key, flights) %>% count(candidate)
```

To get an overview over all tables with primary keys, use `cdm_get_all_pks()`:

```{r}
cdm_get_all_pks(cdm_nycflights13(cycle = TRUE))
```

Here we used the prepared `dm` object `cdm_nycflights13(cycle = TRUE)` as an example. 
This object already has all keys pre-set.

## Foreign keys

Useful functions for managing foreign key relations include:

1. `cdm_add_fk()`
2. `cdm_has_fk()`
3. `cdm_get_fk()`
4. `cdm_rm_fk()`
5. `cdm_enum_fk_candidates()`
6. `cdm_get_all_fks()`

Now it gets (even more) interesting: we want to define relations between different tables.
With the `cdm_add_fk()` function you can define which column of which table points to another table's column.

When dealing with data models, a foreign key from one table has to always point to a primary key of another table.
The primary key of the referred table must be set with `cdm_add_pk()`.
`cdm_add_fk()` will find the primary key column of the referenced table by itself and make the indicated column of the child table point to it.

```{r}
flights_dm_with_key %>% cdm_add_fk(flights, origin, airports)
```
This will throw an error:
```{r error=TRUE}
flights_dm %>% cdm_add_fk(flights, origin, airports)
```
Let's create a `dm` object with a foreign key relation to work with later on:
```{r}
flights_dm_with_fk <- cdm_add_fk(flights_dm_with_key, flights, origin, airports)
```
What if we tried to add another foreign key relation from `flights` to `airports` to the object? Column `dest` might work, since it also contains airport codes:
```{r error=TRUE}
flights_dm_with_fk %>% cdm_add_fk(flights, dest, airports)
```
As you can see, behind the scenes, checks are executed automatically by the functions of `dm` to prevent steps that do not comply with the logic of data models.

Use `cdm_has_fk()` for checking if a foreign key exists that is pointing from one table to another:
```{r}
flights_dm_with_fk %>% cdm_has_fk(flights, planes)
flights_dm_with_fk %>% cdm_has_fk(flights, airports)
```

If you want to access the name of the column which acts as a foreign key of one table to another table's column, use `cdm_get_fk()`:
```{r}
flights_dm_with_fk %>% cdm_get_fk(flights, planes)
flights_dm_with_fk %>% cdm_get_fk(flights, airports)
```

Remove foreign key relations with `cdm_rm_fk()` (parameter `column = NULL` means that all relations will be removed):
```{r error=TRUE}
flights_dm_with_fk %>% 
  cdm_rm_fk(table = flights, column = dest, ref_table = airports) %>% 
  cdm_get_fk(flights, airports)
flights_dm_with_fk %>% 
  cdm_rm_fk(flights, origin, airports) %>% 
  cdm_get_fk(flights, airports)
flights_dm_with_fk %>% 
  cdm_rm_fk(flights, NULL, airports) %>% 
  cdm_get_fk(flights, airports)
```

Since the primary keys are defined in the `dm` object, you do not need to provide the referenced column name of `ref_table`.
This is always the primary key column of the table.

Another function for getting to know your data better (cf. `cdm_enum_pk_candidates()` in ["Primary keys of `dm` objects"](#pk)) is `cdm_enum_fk_candidates()`. Use it to get an overview over foreign key candidates that point from one table to another:

```{r}
cdm_enum_fk_candidates(flights_dm_with_key, weather, airports)
```

Get an overview of all foreign key relations in your data model with`cdm_get_all_fks()`:

```{r}
cdm_get_all_fks(cdm_nycflights13(cycle = TRUE))
```
