---
title: "Filtering in Relational Data Models"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Filtering}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
author: Katharina Brunner
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
fansi::set_knit_hooks(knitr::knit_hooks)
options(crayon.enabled = TRUE, width = 75, cli.width = 75)

knit_print.grViz <- function(x, ...) {
  x %>%
    DiagrammeRsvg::export_svg() %>%
    c("`````{=html}\n", ., "\n`````\n") %>%
    knitr::asis_output()
}
```

The {dm} package offers functions to work with relational data models in R.

This document introduces you to the filtering functions and shows you how to apply them to data that is seperated into multiple tables.

The example is working with data from the {nycflights13} package, that contains tables linked with each other.

First, we load the packages we need:

```{r message=FALSE, warning=FALSE}
library(dm)
library(tidyverse)
library(nycflights13)
```

## Data: nycflights13

To explore filtering with {dm}, we'll use the {nycflights13} data with its tables `flights`, `planes`, `airlines`, `airports` and `weather`.

This dataset contains all 336 776 flights that departed from New York City in 2013 with 3322 different planes and 1458 airports involved.
The data comes from the US Bureau of Transportation Statistics, and is documented in `?nycflights13::flights`.

First of all, we have to create a `dm` object from the {nycflights13} data.
The built-in `dm::cdm_nycflights13()` function takes care of this.

A [data model object](https://krlmlr.github.io/dm/articles/dm-class-and-basic-operations.html#class-dm) contains the data as well as metadata.

If you want to create a `dm` from other tables, please use the `new_dm()`, `dm()` or `as_dm()` functions.

```{r}
dm <- cdm_nycflights13()
```

A `dm` object output with data and metadata consists of three elements: 

1. the table sources,
2. the data model,
3. and a review of active filter conditions.

The console output is colored for a clear output:

```{r}
dm
```

Now we know that there are five tables in our `dm` object.
But how are they connected?
These relations are best displayed as a visualization of the entity-relationship model:

```{r}
cdm_draw(dm)
```

You can look at a single table with `tbl`.
Print the `airports` table:

```{r}
tbl(dm, "airports")
```

## Filtering a `dm` object {#filter}

`cdm_filter()` allows you to select a subset of a `dm` object.

### How it works

Filtering a `dm` object is not that different from filtering a dataframe or tibble with `dplyr::filter()`.

The corresponding {dm} function is `dm::cdm_filter()`. 
With this function one or more filter conditions can be set for one of the `dm` object's tables.
These conditions are not immediately evaluated, but are stored in the object until the moment when they are needed.
There are essentially two ways of causing the filter conditions to take effect:

1. Calling one of `tbl()`, `[[.dm()`, `$.dm()`, i.e. accessing one of the tables of the `dm`. 
In this case, all tables that are connected to the requested table and have filter conditions set are taken into account in the following way:
    - the filter conditions for each of those tables are evaluated (also the requested table can be filtered of course), reducing the numbers of rows of the respective tables
    - filtering semi-joins are successively performed along the paths from each of the filtered to the requested table, each join reducing the left-hand side tables of the joins to only those of their rows with key values that have corresponding values in key columns of the right-hand side tables of the join.
    - eventually the requested table is returned, containing only the the remaining rows after the filtering joins
1. Calling `cdm_apply_filters()` or `compute()` (method for `dm` objects) on a `dm`: this results in a new `dm`, containing the same tables as before. 
Each of the tables is the result of a `tbl()` call, meaning that the effects of the filter conditions on each of the tables are taken into account.

Currently, this only works if the graph induced by the foreign key relations is cycle free.
Luckily, this is the default for `cdm_nycflights13()`.

Mind that several {dm} functions will refuse to work when unevaluated filter conditions exist, such as for example `cdm_enum_fk_candidates()`, `cdm_enum_pk_candidates()`, `cdm_select_tbl()`, `cdm_rename_tbl()`, `cdm_select()`, `cdm_rename()` and `cdm_nrow()`.
In these cases consider applying the filters with `cdm_apply_filters()` first.

It is also important to mention, that `cdm_get_tables()` will NOT evaluate any filter conditions and will just return the original tables.

### Filtering Examples

Let's see it in action:

**We only want the data that is related with John F. Kennedy International Airport.**

```{r}
filtered_dm <- 
  dm %>%
  cdm_filter(airports, name == "John F Kennedy Intl")
filtered_dm
```

The filter expression is listed in the print output.

You can get the numbers of rows of each table with `cdm_nrow()`. 
Before doing this, you need to apply the filters using `cdm_apply_filters()`:

```{r}
rows_per_table <- 
  filtered_dm %>% 
  cdm_apply_filters() %>% 
  cdm_nrow()
rows_per_table
sum(rows_per_table)
```

The total number of rows in the `dm` drops from `r format(sum(cdm_nrow(dm)), big.mark = " ")` to `r format(sum(cdm_nrow(cdm_apply_filters(filtered_dm))), big.mark = " ")` (the only unaffected table is the disconnected `weather` table).

Next example:

**Get a `dm` object containing data for flights from New York to the Dulles International Airport in Washington D.C., abbreviated with `IAD`.**

```{r}
dm %>% 
  cdm_filter(flights, dest == "IAD") %>% 
  cdm_apply_filters() %>% 
  cdm_nrow()
```

Chaining multiple filters on different tables is also supported.

An Example:

**Get all flights from Delta Air Lines which didn't depart from John F. Kennedy International Airport in May 2013.**

```{r}
dm_delta_may <- dm %>%
  cdm_filter(airlines, name == "Delta Air Lines Inc.") %>%
  cdm_filter(airports, name != "John F Kennedy Intl") %>%
  cdm_filter(flights, month == 5)
dm_delta_may
dm_delta_may %>% 
  cdm_apply_filters() %>% 
  cdm_nrow()
```

You can inspect the filtered tables with `tbl`.

In the `airlines` table, Delta is the only remaining carrier:

```{r}
dm_delta_may %>% 
  tbl("airlines")
```

Which planes were used to service these flights?

```{r}
dm_delta_may %>% 
  tbl("planes")
```

And indeed, there are only flights from `month == 5`:

```{r}
dm_delta_may %>% 
  tbl("flights")
```

For comparison, let's review the equivalent manual query for the `flights` in `dplyr` syntax:

```{r}
airlines_filtered <- filter(airlines, name == "Delta Air Lines Inc.")
airports_filtered <- filter(airports, name != "John F Kennedy Intl")
flights %>%
  semi_join(airlines_filtered, by = "carrier") %>%
  semi_join(airports_filtered, by = c("origin" = "faa")) %>%
  filter(month == 5)
```

The {dm} code is leaner, because the foreign key information is encoded in the object.

## SQL statements behind filtering a `dm` object on a database

{dm} is meant to work with relational data models, locally as well as on databases.
In your project the data is probably not stored locally, but in a remote [relational database](https://krlmlr.github.io/dm/articles/dm-introduction-relational-data-models.html#relational-databases) which can be queried with SQL statements.

You can check the queries by using `sql_render()` from the [`dbplyr` package](https://dbplyr.tidyverse.org/).

Example:

**Print the SQL statements for getting all flights from Delta Air Lines which didn't depart from John F. Kennedy International Airport in May 2013 when the data is stored in a sqlite database.**

To show the SQL query behind a `cdm_filter()`, we copy the `flights`, `airlines` and `airports` tables from the `nyflights13` dataset to a temporary in-memory database using the built-in function `cdm_copy_to()` and `dbplyr::src_memdb`.

Then we filter the data and print the corresponding SQL statement with `dbplyr::sql_render()`.

```{r, warning=F}
dm %>%
  cdm_select_tbl(flights, airlines, airports) %>%
  cdm_filter(flights, month == 5) %>% 
  cdm_copy_to(dbplyr::src_memdb(), ., unique_table_names = TRUE) %>% 
  cdm_filter(airlines, name == "Delta Air Lines Inc.") %>%
  cdm_filter(airports, name != "John F Kennedy Intl") %>%
  cdm_apply_filters() %>% 
  cdm_get_tables() %>%
  map(dbplyr::sql_render)
```

Further reading: {dm}'s function for copying data [from and to databases](https://krlmlr.github.io/dm/articles/dm.html#copy).
