---
title: "Low level operations"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Low level operations}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
library(tibble)
library(dm)

knitr::opts_chunk$set(
  error = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

Here we want to present some specialised functions for conducting some basic tests regarding key conditions and generally the relations between tables.
Additionally we will show some functions for splitting and uniting tables.

## Testing key constraints

This section contains information and examples for the functions:

1. `check_key(.data, ...)`
1. `check_if_subset(t1, c1, t2, c2)`
1. `check_set_equality(t1, c1, t2, c2)`

When you have tables (data frames) that are connected by key relations, this package might help you verify the assumed key relations and/or determine existing key relations between the tables. 
For example, if you have tables:

```{r}
data_1 <- tibble(a = c(1, 2, 1), b = c(1, 4, 1), c = c(5, 6, 7))
data_2 <- tibble(a = c(1, 2, 3), b = c(4, 5, 6), c = c(7, 8, 9))
```

and you want to know if `a` is a primary key for `data_1`, you can use the `check_key()` function:

```{r}
check_key(data_1, a)
```

Mind the error message when a test is not passed.

For `data_2`, column `a` is a key:

```{r}
check_key(data_2, a)
```

To see if a column of one table contains only values that are also present in another column of another table, you can use `check_if_subset()`:

```{r}
check_if_subset(data_1, a, data_2, a)
```

This is important to understand if this column is a foreign key to the other table.
What about the inverse relation?

```{r}
check_if_subset(data_2, a, data_1, a)
```

One should keep in mind, that `check_if_subset()` does not test if column `c2` is a unique key of table `t2`.
In order to find out if a (child) table `t1` contains a column `c1` that is a foreign key to a (parent) table `t2` with the corresponding column `c2`, one would use the following approach:

```{r, eval = FALSE}
check_key(t2, c2)
check_if_subset(t1, c1, t2, c2)
```

To check both directions at once - basically answering the questions: are the unique values of `c_1` in `t_1` the same as those of `c_2` in `t_2`? - {dm} provides the function `check_set_equality()`:

```{r}
check_set_equality(data_1, a, data_2, a)
```

Bringing one more table into the game, we can show how it looks, when the test is passed:

```{r}
data_3 <- tibble(a = c(2, 1, 2), b = c(4, 5, 6), c = c(7, 8, 9))

check_set_equality(data_1, a, data_3, a)
```

If the test is passed, the return value of the function will be the first table parameter (invisibly).
This ensures that the functions can be used in a pipe conveniently.

## Testing cardinalities between two tables

This section contains information and examples for the functions:

1. `check_cardinality_0_n(parent_table, primary_key_column, child_table, foreign_key_column)`
1. `check_cardinality_1_n(parent_table, primary_key_column, child_table, foreign_key_column)`
1. `check_cardinality_0_1(parent_table, primary_key_column, child_table, foreign_key_column)`
1. `check_cardinality_1_1(parent_table, primary_key_column, child_table, foreign_key_column)`

The four functions for testing for a specific kind of cardinality of the relation all require a parent table and a child table as input.
All these functions first test if this requirement is fulfilled by checking if: 

1. `primary_key_column` is a unique key for `parent_table`
2. The set of values of `foreign_key_column` is a subset of the set of values of `primary_key_column`

The cardinality specifications `0_n`, `1_n`, `0_1`, `1_1` refer to the expected relation, that the child table has with the parent table.
The numbers '0', '1' and 'n' refer to the number of values in the child table's column (`foreign_key_column`) that correspond to each value of the parent table's column (`primary_key_column`). 
'n' means more than one in this context, with no upper limit.

`0_n` means, that for each value of the `parent_key_column`, the number of corresponding records in the child table is unrestricted.
`1_n` means, that for each value of the `parent_key_column` there is at least one coresponding record in the child table.
This means that there is a "surjective" relation from the child table to the parent table w.r.t. the specified columns, i.e. for each parent table column value there exists at least one equal child table column value.

`0_1` means, that for each value of the `parent_key_column`, minimally '0' and maximally '1' value has to correspond to it in the child table's column. 
This means that there is a "injective" relation from the child table to the parent table w.r.t. the specified columns, i.e. no parent table column value is addressed multiple times. 
But not all of the parent table column values have to be referred to.

`1_1` means, that for each value of the `parent_key_column`, precisely '1' value has to correspond to it in the child table's column. 
This means that there is a "bijective" ("injective" AND "surjective") relation between the child table and the parent table w.r.t. the specified columns, i.e. the set of values of the two columns is equal and there are no duplicates in either of them.

### Examples

Given the following three data frames:
```{r}
d1 <- tibble::tibble(a = 1:5)
d2 <- tibble::tibble(c = c(1:5,5))
d3 <- tibble::tibble(c = 1:4)
```

Here are some examples for the usage of the functions:

```{r}
# This does not pass, `c` is not unique key of d2:
check_cardinality_0_n(d2, c, d1, a)
```


```{r}
# This passes, multiple values in d2$c are allowed:
check_cardinality_0_n(d1, a, d2, c)
```


```{r}
# This does not pass, injectivity is violated:
check_cardinality_1_1(d1, a, d2, c)
```


```{r}
# This passes:
check_cardinality_0_1(d1, a, d3, c)
```

## Table surgery

Relevant functions are:

1. `decompose_table(.data, new_id_column, ...)`
1. `reunite_parent_child(child_table, parent_table, id_column)`
1. `reunite_parent_child_from_list(list_of_parent_child_tables, id_column)`

The first function implements table normalization.
An existing table is split into a parent table (i.e. a lookup table) and a child table (containing the observations), linked by a key column (here: `new_id_column`).
Basically, a foreign key relation would be created, pointing from the child table's `new_id_column` to the parent table's corresponding column, which can be seen as the parent table's primary key column.
The function `decompose_table()` does this, as you can see in the following example:
```{r}
mtcars_tibble <- as_tibble(mtcars)
mtcars_tibble
decomposed_table <- decompose_table(mtcars_tibble, am_gear_carb_id, am, gear, carb)
decomposed_table
```

A new column is created, with which the two tables can be joined by again, essentially creating the original table.

The functions which do the inverse operation, i.e. joining a parent and a child table and subsequently dropping the `new_id_column`, are `reunite_parent_child()` and `reunite_parent_child_from_list()`.
The former takes as arguments two tables and the unquoted name of the ID column, and the latter takes as arguments a list of two tables plus the unquoted name of the ID column:

```{r}
parent_table <- decomposed_table$parent_table
child_table <- decomposed_table$child_table
reunite_parent_child(child_table, parent_table, id_column = am_gear_carb_id)
```

```{r eval = FALSE}
# Shortcut:
reunite_parent_child_from_list(decomposed_table, id_column = am_gear_carb_id)
```

Currently these functions exist only as a low-level operation on tables.
We plan to extend this operation to `dm` objects.
