---
title: "Defining a 'dm' object for your data"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing your data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


This document describes how you can get your data into a `dm` object. 

Although the example that we will be using is included in `dm`, and you can see it by running `dm_nycflights13()`, we will be going through the built-in example one more time here, step by step.

The five tables that we are working with contain information about all flights that departed from the airports of New York to other destinations in the United States in 2013, and are available through the [`nycflights13`](https://github.com/hadley/nycflights13) package: 

- `flights` represents the trips taken by planes 
- `airlines` includes 
  - the names of transport organizations (`name`) 
  - their abbreviated codes (`carrier`)
- `airports` indicates the ports of departure (`origin`) and of destination (`dest`)
- `weather` contains meterological information at each hour
- `planes` describes characteristics of aircrafts

Once we've loaded `nycflights13`, the aforementioned tables are all in our work environment, ready to be accessed.

```{r setup, message=FALSE}
library(dm)
library(nycflights13)
```

# Adding Tables

First, we will tell `dm` which tables we want to work with and how they are connected. 
For that, we can use `dm` or `as_dm()`, which accepts lists of objects. 
You must explicitly name the objects if you use `list()` (e.g., `list("airlines" = airlines, "flights" = flights)`) — here we use `tibble::lst()`, which automatically names the components of the list.

```{r}
flights_dm_no_keys <- tibble::lst(airlines, airports, flights, planes, weather) %>%
  as_dm()
```

# Defining Keys

Even though you now have a new `dm` object that contains all your data, some key details are still missing that determine how your five tables are connected (the foreign keys), and which column(s) uniquely identify the observations (the primary keys).

## Primary Keys

`dm` offers `dm_enum_pk_candidates()` to identify primary keys and `dm_add_pk()` to add them. 

```{r}
dm_enum_pk_candidates(
  dm = flights_dm_no_keys,
  table = planes
)
```

Now, add the primary keys that you have identified:

```{r}
flights_dm_only_pks <- flights_dm_no_keys %>%
  dm_add_pk(table = airlines, column = carrier) %>%
  dm_add_pk(airports, faa) %>%
  dm_add_pk(planes, tailnum)
flights_dm_only_pks
```

To review the primary keys after setting them, call `dm_get_all_pks()`.

```{r}
dm_get_all_pks(flights_dm_only_pks)
```

## Foreign Keys

```{r}
dm_enum_fk_candidates(
  dm = flights_dm_only_pks,
  table = flights,
  ref_table = airlines
)
```

To define how your tables are related, use `dm_add_fk()` to add foreign keys. First, define the tables that you wish to connect by parameterizing the `dm_add_fk()` function with `table` and `ref_table` options. 

Then indicate in `column` which column of `table` refers to `ref_table`'s primary key, which you've defined above. 
Use `check = FALSE` to omit consistency checks.

```{r}
flights_dm_all_keys <- flights_dm_only_pks %>%
  dm_add_fk(table = flights, column = tailnum, ref_table = planes, check = FALSE) %>%
  dm_add_fk(flights, carrier, airlines) %>%
  dm_add_fk(flights, origin, airports)
flights_dm_all_keys
```

## Retrieving Keys

To retrieve your keys later on, use `dm_get_all_fks()`, or `dm_get_fk()` for its singular version. 

```{r}
dm_get_all_pks(flights_dm_all_keys)
```

Voilà, here's your `dm` object that you can work with:

```{r}
flights_dm_all_keys
```
