---
title: "Visualizing 'dm' objects"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing 'dm' objects}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
fansi::set_knit_hooks(knitr::knit_hooks)
options(crayon.enabled = TRUE, width = 75, cli.width = 75)

knit_print.grViz <- function(x, ...) {
  x %>% 
    DiagrammeRsvg::export_svg() %>% 
    c("`````{=html}\n", ., "\n`````\n") %>% 
    knitr::asis_output()
}
```


Once you have all primary keys set and all foreign key relations defined, a graphical representation of your data model offers a condensed view over the tables and relationships.
Relevant functions for visualising the `dm` object's data model are:^[The functions in this section rely heavily on related functions of {datamodelr}.]

1. `dm_draw()`
1. `dm_set_colors()`
1. `dm_get_colors()`
1. `dm_get_available_colors()`

We use the prepared example `dm` object `dm_nycflights13(cycle = TRUE)`:
```{r}
library(dm)
flights_dm_w_many_keys <- dm_nycflights13(color = FALSE)
flights_dm_w_many_keys
```

The schema is drawn with `dm_draw()`. 
```{r}
dm_draw(flights_dm_w_many_keys)
```

You can use colors to visually group your tables into families (which could reflect their logical grouping).
Currently, there are only eight different colors available (plus optional borders around the tables).
To find out which colors exist, use `dm_get_available_colors()`:
```{r}
dm_get_available_colors()
```

The assignment is carried out with `dm_set_colors()`, using a syntax similar to `switch()`.
Any table you don't specifically assign will not be altered. 
The result of `dm_set_colors()` is a `dm` object.
The information about the color is stored in its `data_model`-part.
```{r}
flights_dm_w_many_keys_and_colors <-
  flights_dm_w_many_keys %>% 
  dm_set_colors(
    flights = "blue",
    airlines = ,
    airports = "orange",
    planes = "green_nb"
  )
```

Draw the schema with `dm_draw()`. 
```{r}
dm_draw(flights_dm_w_many_keys_and_colors)
```

The colors can be queried with `dm_get_colors()`. 
```{r}
dm_get_colors(flights_dm_w_many_keys_and_colors)
```

See the documentation for `dm_draw()` for further options.
One important argument is `view_type`.
Besides the default `"keys_only"`, it accepts `"all"` to display all columns, and `"title_only"` to show only the title of the table.

```{r}
flights_dm_w_many_keys_and_colors %>% 
  dm_draw(view_type = "title_only")
```

In case you would like to only see a subset of the whole data model, use `dm_select_tbl()` before drawing:

```{r}
flights_dm_w_many_keys_and_colors %>% 
  dm_select_tbl(flights, airports, planes) %>% 
  dm_draw()
```
