---
title: "Visualizing 'dm' objects"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing 'dm' objects}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
fansi::set_knit_hooks(knitr::knit_hooks)
options(crayon.enabled = TRUE, width = 75, cli.width = 75)

knit_print.grViz <- function(x, ...) {
  x %>% 
    DiagrammeRsvg::export_svg() %>% 
    c("`````{=html}\n", ., "\n`````\n") %>% 
    knitr::asis_output()
}
```

Once you have all your primary keys set and all foreign key relations defined, a graphical representation of your data model offers a condensed view of the tables and the relationships between the tables.
The following functions can be used to visualize the `dm` object:^[The code for the functions in this section is borrowed from the [{datamodelr}](https://github.com/bergant/datamodelr) package.]

1. `cdm_draw()`
1. `cdm_set_colors()`
1. `cdm_get_colors()`
1. `cdm_get_available_colors()`

We use the prepared example `dm` object `cdm_nycflights13(cycle = TRUE)`:
```{r}
library(dm)
flights_dm_w_many_keys <- cdm_nycflights13(color = FALSE)
flights_dm_w_many_keys
```

The schema is drawn with `cdm_draw()`. 
```{r}
cdm_draw(flights_dm_w_many_keys)
```

You can use colors to visually group your tables into families to reflect their logical grouping.
Currently, eight different colors are available, plus optional borders around tables.
To find out which colors exist, use `cdm_get_available_colors()`:
```{r}
cdm_get_available_colors()
```

Colors are assigned with `cdm_set_colors()` using syntax that is similar to `switch()`.
Any table to which you do not specifically assign a color will not be altered. 
The result of `cdm_set_colors()` is a `dm` object.
The information about the color is stored together with the rest of the metadata.
```{r}
flights_dm_w_many_keys_and_colors <-
  flights_dm_w_many_keys %>% 
  cdm_set_colors(
    flights = "blue",
    airlines = ,
    airports = "orange",
    planes = "green_nb"
  )
```

Draw the schema with `cdm_draw()`. 
```{r}
cdm_draw(flights_dm_w_many_keys_and_colors)
```

The colors can be queried with `cdm_get_colors()`. 
```{r}
cdm_get_colors(flights_dm_w_many_keys_and_colors)
```

See the documentation for `cdm_draw()` for further options.
One important argument is `view_type`.
Besides the default `"keys_only"`, it accepts `"all"` to display all columns, and `"title_only"` to show only the title of the table.

```{r}
flights_dm_w_many_keys_and_colors %>% 
  cdm_draw(view_type = "title_only")
```

If you would like to visualize only a some of the tables, use `cdm_select_tbl()` before drawing:

```{r}
flights_dm_w_many_keys_and_colors %>% 
  cdm_select_tbl(flights, airports, planes) %>% 
  cdm_draw()
```
