---
title: "Getting started with dm"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{First read: Getting started with dm}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

``````{r setup, include = FALSE}
source("setup/setup.R")
``````

## Introduction

In this article we're going show you how easy it is to move from connecting to the RDBMS holding your data to producing the results you need.
It's meant to be a quick and friendly introduction to {dm}, so it is low on details and caveats, but links are provided at the end to detailed documnentation.
(If your data is in data frames instead of an RDBMS and you're in a hurry, jump over to `vignette("dm-df")`.)

## Creating a dm object

dm objects can be created from individual tables or loaded directly from a relational data model on an RDBMS.
For this demonstration we're going work with a model hosted on a public server.

The first thing we need is a connection to the RDBMS hosting the data.

``````{r }
library(RMariaDB)

fin_db <- dbConnect(
  MariaDB(),
  username = 'guest',
  password = 'relational',
  dbname = 'Financial_ijs',
  host = 'relational.fit.cvut.cz'
)
``````

We create a dm object from an RDBMS using `dm_from_src()`, passing in the connection object we just created as the first argument.

``````{r message = FALSE}
library(dm)

fin_dm <- dm_from_src(fin_db)
fin_dm
``````

The dm object interrogates the RDBMS for table and column information and, where implemented, primary and foreign keys.
Currently, primary and foreign keys are only available from Postgres and SQL Server.

## Linking tables by adding keys

The public RDBMS we've connected to is a MariaDB, which does not provide key metadata.
That's not a problem.
The lack of keys is easily fixed.
dm allows us to define primary and foreign keys ourselves.

```{r keys}

fin_dm_keys <-
  fin_dm %>%
  dm_add_pk(accounts, id) %>%
  dm_add_pk(loans, id) %>%
  dm_add_fk(loans, account_id, accounts) %>%
  dm_add_pk(trans, id) %>%
  dm_add_fk(trans, account_id, accounts) %>%
  dm_add_pk(districts, id) %>%
  dm_add_fk(accounts, district_id, districts)
```

## Visualizing a data model

Having a diagram of the data model is the quickest way to verify we're on the right track.
We can display a visual summary of the dm at any time.
The default is to display the table name, any defined keys and their links to other tables.

Visualizing the dm in its current state we see the keys we have created and how they link the tables together.

```{r visualize_keys}

fin_dm_keys %>%
  dm_set_colors(green = loans, orange = accounts, red = trans, grey = districts) %>%
  dm_draw()

```

## Accessing a data model as a table

If we want to perform modeling or analysis on this relational model we need to transform it into a tabular format that R functions can work with.
`dm_flatten_to_tbl()` will automatically follow foreign keys across tables to gather all the available columns into a single table.

```{r flatten}

fin_dm_keys %>%
  dm_flatten_to_tbl(loans)

#FIXME maybe a show a simple operation on the table once it has been extracted?

```

## Operations on table data within a dm

We don't need to take the extra step of exporting the data to work with it.
Through the dm object we have complete access to dplyr's data manipulation verbs.
These operate on the data within individual tables.

To work with a particular table we use `dm_zoom_to()` to set the context to our chosen table.
Then we can perform any of the operations we want.

```{r zoom}
fin_dm_total <-
  fin_dm_keys %>%
  dm_zoom_to(loans) %>%
  group_by(account_id) %>%
  summarize(total_amount = sum(amount, na.rm = TRUE)) %>%
  ungroup() %>%
  dm_insert_zoomed("total_loans")

fin_dm_total$total_loans

```

Note that in the above example we use `dm_insert_zoomed()` to add the results as a new table to our data model.
This table is temporary and will be deleted when our session ends.
If you want to make permanent changes to your data model on an RDBMS please see the documentation on [persistence](vignettes/dm-db#persist #FIXME - link to anchor?).

## Checking constraints

It's always smart to check that your data model follows its specifications.
When building our own model or changing existing models by adding tables or keys, it is even more important that the new model is validated.

`dm_examine_constrains()` checks all primary and foreign keys and reports if they violate their expected constraints.

```{r constraints}
fin_dm_total %>%
  dm_examine_constraints()
```

For more on constraint checking, including cardinality, finding candidate columns for keys, and normalization, see the `vignette("dm-low-level")` documentation.

## Next Steps

Now that you have been introduced to the basic operation of dm, the next step is to learn more about the dm methods that your particular use case requires.

Is your data in an RDBMS? Then move on to `vignette("dm-db")` for a more detailed looking at working with an existing relational data model.

If your data is in data frames, then you want to read `vignette("dm-df")` next.

If you feel you need to know more about relational data models in order to get the most out of dm, check out `vignette("dm-introduction-relational-data-models")`.

If you're familiar with relational data models but want to know how to work with them in dm, then any of `vignette("dm-joining")`, `vignette("dm-filtering")`, or `vignette("dm-zoom-to-table")` is a good next step.

## Further reading

`vignette("dm-db")` - This article covers accessing and working with RDBMSs within your R session, including manipulating data, filling in missing relationships between tables, getting data out of the RDBMS and into your model, and deploying your data model to an RDBMS.

`vignette("dm-df")` - Is your data in local data frame? This article covers creating a data model from your local data frames, including building the relationships in your data model, verifying your model, and leveraging the power of dplyr to operate on your data model.

`vignette("dm-zoom-to-table")` - Learn how to manipulate the data in the tables in your dm using dplyr verbs.

`vignette("dm-filtering")` - Not familiar with SQL? This introduction to dm's filtering methods will show you how to apply them to data that is separated into multiple tables. Handy for when you want to work with a subset of the data on an RDBMS.

`vignette("dm-visualization")` - A quick guide to dm's methods for drawing your data model for exploration, verification or documentation purposes. Learn how to customize the diagrams.

`vignette("dm-low-level")` - If your data model needs to be rock solid, this article covers the dm methods for checking every aspect of the model. It covers key constraints, cardinalities between tables, and methods for normalization as you construct your data model.

`vignette("dm-introduction-relational-data-models")` - Do you know all about data frames but very little about relational data models? This quick introduction will walk you through the key similarities and differences, and show you how to move from individual data frames to a relational data model.

`vignette("dm-joining")` - Joining is how separate, linked tables are virtually combined in order to perform operations, generate summaries, or prepare data for extraction. This article covers dm's methods for creating the different kind of joins along with usage examples.

`vignette("dm-class-and-basic-operations")` - #FIXME do we still need this? Could it be re-written? More of an outline of the class, a list of methods, then its operation: creating dm objects, immutability, pipelines?

## Standing on the shoulders of giants

The {dm} package follows the tidyverse principles:

- `dm` objects are immutable (your data will never be overwritten in place)
- many functions used on `dm` objects are pipeable (i.e., return new `dm` objects)
- tidy evaluation is used (unquoted function parameters are supported)

The {dm} package builds heavily upon the [{datamodelr} package](https://github.com/bergant/datamodelr), and upon the [tidyverse](https://www.tidyverse.org/).
We're looking forward to a good collaboration!

The [{polyply} package](https://github.com/russHyde/polyply) has a similar intent with a slightly different interface.

The [{data.cube} package](https://github.com/jangorecki/data.cube) has quite the same intent using `array`-like interface.

Articles in the [{rquery} package](https://github.com/WinVector/rquery) discuss [join controllers](https://github.com/WinVector/rquery/blob/master/extras/JoinController.md) and [join dependency sorting](https://github.com/WinVector/rquery/blob/master/extras/DependencySorting.md), with the intent to move the declaration of table relationships from code to data.

The [{tidygraph} package](https://github.com/thomasp85/tidygraph) stores a network as two related tables of `nodes` and `edges`, compatible with {dplyr} workflows.

In object-oriented programming languages, [object-relational mapping](https://en.wikipedia.org/wiki/Object-relational_mapping) is a similar concept that attempts to map a set of related tables to a class hierarchy.

