---
title: "Getting started with dm"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{First read: Getting started with dm}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

``````{r setup, include = FALSE}
source("setup/setup.R")
``````

## Introduction

In this article we're going show you how easy it is to move from connecting to the database holding your data to producing the results you need.
It's meant to be a quick and friendly introduction to {dm}, so it is low on details and caveats.
Links to detailed documentation are provided at the end.
(If your data is in data frames instead of a database and you're in a hurry, jump over to `vignette("howto-dm-df")`.)

## Creating a dm object

dm objects can be created from individual tables or loaded directly from a relational data model on an RDBMS (relational database management system).

For this demonstration, we're going to work with a model hosted on a public server.
The first thing we need is a connection to the RDBMS hosting the data.

``````{r connect, eval = FALSE}
library(RMariaDB)

fin_db <- dbConnect(
  MariaDB(),
  username = "guest",
  password = "relational",
  dbname = "Financial_ijs",
  host = "relational.fit.cvut.cz"
)
``````

``````{r connect-real, show = FALSE}
library(RMariaDB)

fin_db <- dm:::financial_db_con()
``````

We create a dm object from an RDBMS using `dm_from_con()`, passing in the connection object we just created as the first argument.

``````{r load-full}
library(dm)

fin_dm <- dm_from_con(fin_db)
fin_dm
``````

The dm object interrogates the RDBMS for table and column information, and primary and foreign keys.


## Selecting tables

The dm object can be accessed like a named list of tables:

```{r names}
names(fin_dm)
fin_dm$loans
dplyr::count(fin_dm$trans)
```

Additionally, most `dm` functions are [pipe-friendly](https://r4ds.had.co.nz/pipes.html) and support [tidy evaluation](https://adv-r.hadley.nz/metaprogramming.html).
We can use `[` or the `dm_select_tbl()` verb to derive a smaller dm with the `loans`, `accounts`, `districts` and `trans` tables:

```{r select}
fin_dm_small <- fin_dm[c("loans", "accounts", "districts", "trans")]
fin_dm_small <-
  fin_dm %>%
  dm_select_tbl(loans, accounts, districts, trans)
```


## Linking tables by adding keys

In many cases, `dm_from_con()` already returns a dm with all keys set.
If not, dm allows us to define primary and foreign keys ourselves.
For this, we use `learn_keys = FALSE` to obtain a `dm` object with only the tables.

``````{r load}
library(dm)

fin_dm_small <- 
  dm_from_con(fin_db, learn_keys = FALSE) %>% 
  dm_select_tbl(loans, accounts, districts, trans)
``````

In our data model, `id` columns uniquely identify records in the `accounts` and `loans` tables, and can be used as a primary key.
A primary key is defined with `dm_add_pk()`.
Each loan is linked to one account via the `account_id` column in the `loans` table, the relationship is established with `dm_add_fk()`.

```{r keys}
fin_dm_keys <-
  fin_dm_small %>%
  dm_add_pk(table = accounts, columns = id) %>%
  dm_add_pk(loans, id) %>%
  dm_add_fk(table = loans, columns = account_id, ref_table = accounts) %>%
  dm_add_pk(trans, id) %>%
  dm_add_fk(trans, account_id, accounts) %>%
  dm_add_pk(districts, id) %>%
  dm_add_fk(accounts, district_id, districts)
```

## Visualizing a data model

Having a diagram of the data model is the quickest way to verify we're on the right track.
We can display a visual summary of the dm at any time.
The default is to display the table name, any defined keys, and their links to other tables.

Visualizing the dm in its current state, we can see the keys we have created and how they link the tables together.
Color guides the eye.

```{r visualize_keys}
fin_dm_keys %>%
  dm_set_colors(green = c(loans, accounts), darkblue = trans, grey = districts) %>%
  dm_draw()
```

## Accessing a data model as a table

If we want to perform modeling or analysis on this relational model, we need to transform it into a tabular format that R functions can work with.
`dm_squash_to_tbl()` will automatically follow foreign keys across tables to gather all the available columns into a single table.

```{r squash}
fin_dm_keys %>%
  dm_squash_to_tbl(loans)
```

Apart from the rows printed above, no data has been fetched from the database.
Use `select()` to reduce the number of columns fetched, and `collect()` to retrieve the entire result for local processing.

```{r model}
loans_df <-
  fin_dm_keys %>%
  dm_squash_to_tbl(loans) %>%
  select(id, amount, duration, A3) %>%
  collect()

model <- lm(amount ~ duration + A3, data = loans_df)

model
```


## Operations on table data within a dm

We don't need to take the extra step of exporting the data to work with it.
Through the dm object, we have complete access to dplyr's data manipulation verbs.
These operate on the data within individual tables.

To work with a particular table we use `dm_zoom_to()` to set the context to our chosen table.
Then we can perform any of the dplyr operations we want.

```{r zoom}
fin_dm_total <-
  fin_dm_keys %>%
  dm_zoom_to(loans) %>%
  group_by(account_id) %>%
  summarize(total_amount = sum(amount, na.rm = TRUE)) %>%
  ungroup() %>%
  dm_insert_zoomed("total_loans")

fin_dm_total$total_loans
```

Note that, in the above example, we use `dm_insert_zoomed()` to add the results as a new table to our data model.
This table is temporary and will be deleted when our session ends.
If you want to make permanent changes to your data model on an RDBMS, please see the "Persisting results" section in `vignette("howto-dm-db")`.


## Checking constraints

It's always smart to check that your data model follows its specifications.
When building our own model or changing existing models by adding tables or keys, it is even more important that the new model is validated.

`dm_examine_constrains()` checks all primary and foreign keys and reports if they violate their expected constraints.

```{r constraints}
fin_dm_total %>%
  dm_examine_constraints()
```

For more on constraint checking, including cardinality, finding candidate columns for keys, and normalization, see `vignette("tech-dm-low-level")`.

``````{r disconnect, echo = FALSE, results = "hide"}
dbDisconnect(fin_db)
``````

## Next Steps

Now that you have been introduced to the basic operation of dm, the next step is to learn more about the dm methods that your particular use case requires.

Is your data in an RDBMS? Then move on to `vignette("howto-dm-db")` for a more detailed looking at working with an existing relational data model.

If your data is in data frames, then you want to read `vignette("howto-dm-df")` next.

If you feel you need to know more about relational data models in order to get the most out of dm, check out `vignette("howto-dm-theory")`.

If you're familiar with relational data models, but want to know how to work with them in dm, then any of `vignette("tech-dm-join")`, `vignette("tech-dm-filter")`, or `vignette("tech-dm-zoom")` is a good next step.

## Further reading

`vignette("howto-dm-db")` -- This article covers accessing and working with RDBMSs within your R session, including manipulating data, filling in missing relationships between tables, getting data out of the RDBMS and into your model, and deploying your data model to an RDBMS.

`vignette("howto-dm-df")` -- Is your data in local data frames? This article covers creating a data model from your local data frames, including building the relationships in your data model, verifying your model, and leveraging the power of dplyr to operate on your data model.

`vignette("howto-dm-theory")` -- Do you know all about data frames but very little about relational data models? This quick introduction will walk you through the key similarities and differences, and show you how to move from individual data frames to a relational data model.

`vignette("tech-dm-join")` -- Joining is how separate, linked tables are virtually combined in order to perform operations, generate summaries, or prepare data for extraction. This article covers dm's methods for creating the different kind of joins along with usage examples.

`vignette("tech-dm-zoom")` -- Learn how to manipulate the data in the tables in your dm using dplyr verbs.

`vignette("tech-dm-filter")` -- Need a subset of your data across tables? This introduction to dm's filtering methods will show you how to apply them to data that is separated into multiple tables.

`vignette("tech-dm-draw")` -- A quick guide to dm's methods for drawing your data model for exploration, verification or documentation purposes. Learn how to customize the diagrams.

`vignette("tech-dm-low-level")` -- If your data model needs to be rock solid, this article covers the dm methods for checking every aspect of the model. It covers key constraints, cardinalities between tables, and methods for normalization as you construct your data model.

`vignette("tech-dm-class")` -- A detailed overview of the `"dm"` class.

`vignette("tech-dm-naming")` -- Making sense of dm's API.

`vignette("tech-dm-cdm")` -- If you're an early adopter, used dm version 0.0.5 or lower and want to migrate your code, this is for you.

## Standing on the shoulders of giants

The {dm} package follows the tidyverse principles:

- `dm` objects are immutable (your data will never be overwritten in place)
- most functions used on `dm` objects are pipeable (i.e., return new `dm` or table objects)
- tidy evaluation is used (unquoted function arguments are supported)

The {dm} package builds heavily upon the [{datamodelr} package](https://github.com/bergant/datamodelr), and upon the [tidyverse](https://www.tidyverse.org/).
We're looking forward to a good collaboration!

The [{polyply} package](https://github.com/russHyde/polyply) has a similar intent with a slightly different interface.

The [{data.cube} package](https://github.com/jangorecki/data.cube) has quite the same intent using `array`-like interface.

Articles in the [{rquery} package](https://github.com/WinVector/rquery) discuss [join controllers](https://github.com/WinVector/rquery/blob/master/extras/JoinController.md) and [join dependency sorting](https://github.com/WinVector/rquery/blob/master/extras/DependencySorting.md), with the intent to move the declaration of table relationships from code to data.

The [{tidygraph} package](https://github.com/thomasp85/tidygraph) stores a network as two related tables of `nodes` and `edges`, compatible with {dplyr} workflows.

In object-oriented programming languages, [object-relational mapping](https://en.wikipedia.org/wiki/Object-relational_mapping) is a similar concept that attempts to map a set of related tables to a class hierarchy.
