---
title: "Persistent row-level modifications of a dm"
date: "`r Sys.Date()`"
author: James Wondrasek
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{How to: Create a dm object from a database}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---


``````{r setup, include = FALSE}
source("setup/setup.R")
``````

## Introduction {#intro}

This tutorial introduces the methods {dm} provides for modifying the data in the tables of a relational model.
There are 6 methods:

* [`dm_rows_insert()`](#insert) - adds new rows
* [`dm_rows_update()`](#update) - changes values in rows
* [`dm_rows_upsert()`](#upsert) - adds new rows or changes values if pre-existing
* [`dm_rows_patch()`](#patch) - fills in missing values
* [`dm_rows_delete()`](#delete) - deletes rows
* [`dm_rows_truncate()`](#truncate) - removes all rows from table, leaving table structure intact

## The dm_rows_* process

All six methods take the same arguments and using them follows the same process. 
To start, a dm object is created containing the tables, and rows, that you want to change.
This dm is then copied into the same source as the dm you want to modify.
With the dm in the same RDBMS as the destination dm, you call the appropriate method, such as `dm_rows_insert()`, to make your planned changes, along with an argument of `in_place = FALSE` so you can confirm you achieve the changes that you want.

This verification can be done visually, looking at row counts and the like, or using {dm}'s constraint checking method, `dm_examine_constraints()`.
The biggest danger is damaging key relations between data spread across multiple tables by deleting or duplicating rows and their keys.
`dm_examine_constraints()` will catch errors where primary keys are duplicated or foreign keys do not have a matching primary key (unless the foreign key value is `NA`).

With the changes confirmed, you execute the method again, this time with the argument `in_place = TRUE` to make the changes permanent.

Each method has its own requirements in order to maintain database consistency. 
These involve constraints on primary key values as they are how rows are identified.

| Method | Requirements |
|--------|--------------|
| dm_rows_insert() | The primary keys must differ from existing records.|
| dm_rows_update() | Primary keys must match for all records to be updated.|
| dm_rows_patch() | Updates missing values in existing records. Primary keys must match for all records to be patched.|
| dm_rows_upsert() | Updates existing records and adds new records, based on the primary key.|
| dm_rows_delete() | Removes matching records based on the primary key. The order in which the tables are processed is reversed.|
| dm_rows_truncate() | Removes all records, only for tables in dm. Truncation is performed by truncating child tables before parent tables. |

For `dm_rows_delete()` and `dm_rows_truncate()`, tables linked to other tables via foreign keys are processed first. 
This is to ensure the integrity of all relations during the deletion or truncation process.
For more details on this see `vignette("howto-dm-db")`.

## Usage {#usage}

To demonstrate the use of these table modifying methods we will create a simple dm object with two tables linked by a foreign key.

Note the foreign key of `NA` in the `child` tibble.
{dm} doesn't check your key values when you create a dm.
You may want to use `dm_examine_constraints()` after it is created, but
be aware that when using `dm_examine_constraints()` NULL (`NA`) foreign keys are allowed and will be counted as a match. 
In some cases this doesn't make sense and non-NULL columns should be enforced by the RDBMS. 
Currently {dm} does not specify or check non-NULL constraints for columns.

``````{r }
library(tidyverse)
library(dm)

parent <- tibble(value = c("A", "B", "C"), pk = 1:3)
child <- tibble(value = c("a", "b", "c"), pk = 1:3, fk = c(1, 1, NA))

demo_dm <-
  dm(parent = parent, child = child) %>%
  dm_add_pk(parent, pk) %>%
  dm_add_pk(child, pk) %>%
  dm_add_fk(child, fk, parent)

demo_dm
dm_examine_constraints(demo_dm)
``````

Table modification methods work as expected on local dm objects, returning a new dm with the modifications in place.
dm objects whose source is a database treat all operations as temporary unless you request explicitly to commit them by including the argument `temporary = FALSE`. 

Below we copy `demo_dm` into an SQLite database.
Note: the default for the method used, `copy_dm_to()`, is to create temporary tables that will be automatically deleted when your sessions ends.
As `demo_dm` will be the destination dm for the examples, the argument `temporary = FALSE` is used to make this distinction apparent.


``````{r }
library(DBI)
sqlite_db <- DBI::dbConnect(RSQLite::SQLite(":memory:"))
demo_sql <- copy_dm_to(sqlite_db, demo_dm, temporary = FALSE)
demo_sql
``````

{dm}'s table modification methods can be piped together to create a repeatable sequence of operations that returns a dm incorporating all the changes required.
This is a common use case for {dm} -- building by hand a sequence of operations using temporary results until it is complete and correct, then committing the result.

## `dm_rows_insert()` {#insert}

To demonstrate `dm_rows_insert()` we create a dm with tables containing the rows to insert and copy it to `sqlite_db`, the same source as `demo_sql`.
For all of the `dm_rows_*` methods the source and destination dm objects must be in the same RDBMS.
You will get an error message if this is not the case.

The code below adds `parent` and `child` table entries for the letter "D".

``````{r }
new_parent <- tibble(value = "D", pk = 4)
new_child <- tibble(value = "d", pk = 4, fk = 4)
dm_insert_in <-
  dm(parent = new_parent, child = new_child) %>%
  copy_dm_to(sqlite_db, ., temporary = TRUE)
dm_insert_out <-
  demo_sql %>%
  dm_rows_insert(dm_insert_in)
dm_insert_out$child
demo_sql$child
``````

This gives us a warning that changes will not be persisted, and inspecting the `child` table of the resulting `dm_insert_out` and `demo_sql`, we can see that's exactly what happened.
{dm} returned to us a dm object with our inserted rows in place, but the underlying database has not changed.

We repeat the operation, this time with the argument `in_place = TRUE` and the changes now persist in `demo_sql`.

``````{r }
dm_insert_out <-
  demo_sql %>%
  dm_rows_insert(dm_insert_in, in_place = TRUE)
demo_sql$child
``````

## `dm_rows_update()` {#update}

`dm_rows_update()` works the same as `dm_rows_insert()`.
We create the dm object and copy it to the same source as the destination.
Here we will change the foreign key for the row in `child` containing "b" to point to the correct row in `parent`.
And we will persist the changes.

``````{r }
new_child <- tibble(value = "b", pk = 2, fk = 2)

dm_update_in <-
  dm(child = new_child) %>%
  copy_dm_to(sqlite_db, ., temporary = TRUE)

dm_update_out <-
  demo_sql %>%
  dm_rows_update(dm_update_in, in_place = TRUE)
demo_sql$child
``````

## `dm_rows_delete()` {#delete}

`dm_rows_delete()` is not currently implemented to work with an RDBMS, so we will shift our demonstrations back to the local R environment.
We've made changes to `demo_sql` so we use `collect()` to copy the current tables out of SQLite.
Note that persistence is not a concern with local dm objects. 
Every operation returns a new dm object containing the changes made.

``````{r }
local_dm <- collect(demo_sql)
local_dm$parent
local_dm$child

dm_deleted <-
  dm(parent = new_parent, child = new_child) %>%
  dm_rows_delete(local_dm, .)
dm_deleted$child
``````

## `dm_rows_patch()` {#patch}

`dm_rows_patch()` updates missing values in existing records.
We use it here to fix the missing foreign key in the `child` table.

``````{r }
updated_child <- tibble(value = "c", pk = 3, fk = 3)

dm_patched <-
  dm(child = updated_child) %>%
  dm_rows_patch(dm_deleted, .)
dm_patched$child
``````

## `dm_rows_upsert()` {#upsert}

`dm_rows_upsert()` updates rows with supplied values if they exist or inserts the supplied values as new rows if they don't.
In this example we add the letter "D" back to our dm, and update the foreign key for "b".

``````{r }
upserted_parent <- tibble(value = "D", pk = 4)
upserted_child <- tibble(value = c("b", "d"), pk = c(2, 4), fk = c(3, 4))

dm_upserted <-
  dm(parent = upserted_parent, child = upserted_child) %>%
  dm_rows_upsert(dm_patched, .)
dm_upserted$parent
dm_upserted$child
``````

## `dm_rows_truncate()` {#truncate}

`dm_rows_truncate()` deletes all the rows in a table while leaving all other related information intact, including column names, column types, and key relations. 
The function derives its name from the SQL `TRUNCATE TABLE` statement, so we will return to our SQLite database to demonstrate its use. 

The first step is to create a dm object with the tables we want to truncate.
`dm_rows_truncate()` looks at the table names in the source dm to select tables for truncation, and then uses their column names to confirm the tables match and should be truncated.

To demonstrate `dm_rows_truncate()` we create a dm with the table we want to truncate.
A convenient way to achieve this is to subset the table from the target dm with `dm_select_tbl()`.
The contents of the table do not matter.

``````{r }
dm_trunc_in <-
  demo_sql %>%
  dm_select_tbl(child)
dm_trunc_in

dm_trunc_out <-
  demo_sql %>%
  dm_rows_truncate(dm_trunc_in, in_place = FALSE)
dm_trunc_out$child
``````

The method `dm_ptype()` returns the prototype of a dm as local object.
This prototype contains all the tables with column names and types, and also keys, but the tables are empty tibbles.
This is a good starting point to create a dm object to be passed to `dm_rows_truncate()` or to create an empty copy of a relational data model to populate.

In the example below, an empty copy of `demo_sql` is created.
Then, rather than selecting the table to truncate, the table that shouldn't be truncated is removed, before copying the dm into the same RDBMS as `demo_sql`.

``````{r }

empty_dm <- dm_ptype(demo_sql)
empty_dm
dm_nrow(empty_dm)

dm_trunc_in <- empty_dm %>% dm_rm_tbl(parent) %>% copy_dm_to(sqlite_db, ., temporary = TRUE)

dm_trunc_out <-
  demo_sql %>%
  dm_rows_truncate(dm_trunc_in, in_place = FALSE)
dm_trunc_out$parent
dm_trunc_out$child

``````


## Conclusion {#conclusion}

The `dm_rows_*` methods give you row-level granularity over the modifications you need to make to your relational model. 
By using the `in_place` argument they all share you can construct and verify your modifications before committing them. 
There are a few limitations, as mentioned in the tutorial, but these will be addressed in future updates to {dm}.

## Next Steps

If this tutorial answered some questions, but opened others, these resources might be of assistance.

Is your data in an RDBMS? `vignette("howto-dm-db")` offers a detailed look at working with an existing relational data model.

If your data is in data frames, then you may want to read `vignette("howto-dm-df")` next.

If you feel you need to know more about relational data models in order to get the most out of dm, check out `vignette("howto-dm-theory")`.

If you're familiar with relational data models but want to know how to work with them in dm, then any of `vignette("tech-dm-join")`, `vignette("tech-dm-filter")`, or `vignette("tech-dm-zoom")` is a good next step.
