---
title: "Loading data into a dm"
date: "`r Sys.Date()`"
author: James Wondrasek, Kirill Müller
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{How to: Create a dm object from a database}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---


``````{r setup, include = FALSE}
source("setup/setup.R")
``````

# Loading data into a dm from a database

A dm object can be created from any database that has a {[DBI](https://dbi.r-dbi.org/)} backend ([see list](https://github.com/r-dbi/backends)).
When a dm object is created it can either import all the tables in the database, the active schema or a limited set.
For some RDBMS, such as Postgres and SQL Server, primary and foreign keys are also imported and do not have to be manually added afterwards.

To demonstrate, we connect to a [relational dataset repository](https://relational.fit.cvut.cz/) with a database server that is publicly accessible without registration.
There is a [financial dataset](https://relational.fit.cvut.cz/dataset/Financial) that contains loan data along with relevant account information and transactions.
We chose this dataset because the relationships between `loan`, `account`, and `transactions` tables are a good representation of databases that record real-world business transactions.
The repository uses a MariaDB server for which {dm} does not currently import primary or foreign keys, so we will need to add them.

``````{r }
library(RMariaDB)
my_db <- dbConnect(
  MariaDB(),
  username = 'guest',
  password = 'relational',
  dbname = 'Financial_ijs',
  host = 'relational.fit.cvut.cz'
)
``````

Creating a dm object takes a single call to `dm_from_src()` with the DBI connection object as its argument.

``````{r message = FALSE}
library(dm)

my_dm <- dm_from_src(my_db)
my_dm
``````

The components of the `my_dm` object are lazy tables powered by {[dbplyr](https://dbplyr.tidyverse.org/)}.
{dbplyr} translates the {[dplyr](https://dplyr.tidyverse.org/)} grammar of data manipulation into queries the database server understands.
Lazy tables defer downloading of table data until results are required for printing or local processing.

## Building a dm from a subset of tables

A dm can also be constructed from individual tables or views.
This is useful for when you want to work with a subset of a database's tables, perhaps from different schemas.

Below we use the `tbl()` function from {dplyr} to extract two tables from the financial database.
Then we create our dm by passing the tables in as arguments.
Note that the tables arguments have to all be from the same source.

``````{r }
dbListTables(my_db)

library(dbplyr)
loans <- tbl(my_db, "loans")
accounts <- tbl(my_db, "accounts")

my_manual_dm <- dm(loans, accounts)
my_manual_dm
``````

## Define Primary and Foreign Keys

Primary keys and foreign keys are how relational database tables are linked with each other.
A primary key is a column that has a unique value for each row within a table.
A foreign key is a column containing the primary key for a row in another table.[^compound]
Foreign keys act as cross references between tables.
They specify the relationships that gives us the *relational* database.

[^compound]: Support for compound keys (consisting of multiple columns) is [planned](https://github.com/krlmlr/dm/issues/3).

The [model diagram](https://relational.fit.cvut.cz/assets/img/datasets-generated/financial.svg) provided by our test database loosely illustrates the intended relationships between tables.
In the diagram we can see that the `loans` table should be linked to the `accounts` table.
Below we create those links in 3 steps:

1. Add a primary key `id` to the `accounts` table
1. Add a primary key `id` to the `loans` table
1. Add a foreign key `account_id` to the `loans` table referencing the `accounts` table

Then we assign colors to the tables and draw the structure of the dm.

Note that when the foreign key is created the primary key in the referenced table does not need to be specified, but the primary key must already be defined.
And, as mentioned above, primary and foreign key constraints on the database are currently only imported for Postgres and SQL Server databases, and only when `dm_from_src()` is used.
This process of key definition needs to be done manually for other databases.

``````{r }
my_dm_keys <-
  my_manual_dm %>%
  dm_add_pk(accounts, id) %>%
  dm_add_pk(loans, id) %>%
  dm_add_fk(loans, account_id, accounts) %>%
  dm_set_colors(green = loans, orange = accounts)

my_dm_keys %>%
  dm_draw()
``````

Once you have instantiated a dm object you can continue to add tables to it.
For tables from the original source for the dm, use `dm_add_tbl()`

``````{r }
trans <- tbl(my_db, "trans")

my_dm_keys %>%
  dm_add_tbl(trans)
``````

For tables from other sources or from the local environment `dplyr::copy_to()` is used.
`copy_to()` is discussed later in this article.

## Transient nature of operations

Like other R objects, a dm is immutable and all operations performed on it are transient unless stored in a new variable.

``````{r }
my_dm_keys

my_dm_trans <-
  my_dm_keys %>%
  dm_add_tbl(trans)

my_dm_trans
``````

And, like {dbplyr}, results are never written to a database unless explicitly requested.

``````{r }
my_dm_keys %>%
  dm_flatten_to_tbl(loans)

my_dm_keys %>%
  dm_flatten_to_tbl(loans) %>%
  sql_render()
``````

## Persisting results

dm provides many of the {dplyr} methods for making changes you have made to the database permanent.
For example, `compute()` forces the computation of the query, or multiple queries, created by the dm to represent all operations that have been performed but not yet evaluated.
The results of `compute()` are stored in temporary tables on the database server and will be deleted when your session ends.

For more details on `compute()` as well as methods and techniques for persisting results please read `vignette('howto-dm-copy`).
You may also wish to read the `compute()` documentation for more details.


## Conclusion

In this tutorial we have demonstrated how simple it is to load a database into a dm object and begin working with it.
While loading a dm from most RDBMS currently requires you to manually set key relations, dm provides methods to make this straightforward.
It is planned that future versions of dm will support automatic key creation for more RDBMS.

The next step is to read `vignette('howto-dm-copy`), where moving your dm into an RDBMS is covered.

## Further reading

`vignette("howto-dm-db")` - This article covers accessing and working with RDBMSs within your R session, including manipulating data, filling in missing relationships between tables, getting data out of the RDBMS and into your model, and deploying your data model to an RDBMS.

`vignette("howto-dm-df")` - Is your data in local data frames? This article covers creating a data model from your local data frames, including building the relationships in your data model, verifying your model, and leveraging the power of dplyr to operate on your data model.

`vignette("howto-dm-theory")` - Do you know all about data frames but very little about relational data models? This quick introduction will walk you through the key similarities and differences, and show you how to move from individual data frames to a relational data model.

#FIXME is this section below unnecessary and can be deleted or do you want it brought back into this tutorial?

## Performing operations on tables by "zooming"

As the dm is a collection of tables, if we wish to perform operations on an individual table we set it as the context for those operations using `dm_zoom_to()`.
See `vignette("tech-dm-zoom")` for more detail on zooming.

dm operations are transient unless persistence is explicitly requested.
To make our chain of manipulations on the selected table permanent we assign the result of `dm_insert_zoomed()` to a new object, `my_dm_total`.
This is a new dm object, derived from `my_dm_keys`, with a new lazy table `total_loans` linked to the `accounts` table.
The subsequent section describes how to materialize the results on the database.

``````{r }
my_dm_total <-
  my_dm_keys %>%
  dm_zoom_to(loans) %>%
  group_by(account_id) %>%
  summarize(total_amount = sum(amount, na.rm = TRUE)) %>%
  ungroup() %>%
  dm_insert_zoomed("total_loans")

my_dm_total$total_loans

my_dm_total %>%
  dm_draw()

my_dm_total$total_loans %>%
  sql_render()
``````

