---
title: "Preparing Your Data"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing your data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

This document describes how you can get your data into a `dm` object. 
Though this example is also built in already — just run `cdm_nycflights13()` — here, we'll go through the process step by step.

The five tables we are working with contain information on flights from airports of the North-American city of New York to other destinations in the United States in 2013 and are available through the package [`nycflights13`](https://github.com/hadley/nycflights13)). 

- `flights` represents the journeys taken by planes, 
- `airlines` includes 
  - transport organizations' names (`name`) 
  - their shortcuts (`carrier`)
- `airports` indicates the ports of departure (`origin`) and of destination (`dest`)
- `weater` the meterological context at each hour
- `planes` describes aircrafts' characteristics

Once we've loaded `nycflights13`, the aforementioned tables are all in our work environment, ready for being accessed.

```{r setup message=FALSE}
library(dm)
library(nycflights13)
```

# Adding Tables

First, we will tell `dm` what tables we want to work with and how they're connected. 
For this, we can use `dm` or `as_dm()`. 
`as_dm()` accepts lists of objects. 
You must explicitly name the objects if you use `list()` (f.e. `list("airlines" = airlines, "flights" = flights)`) — here we use `tibble::lst()` which automatically names the list components.

```{r}
flights_dm_no_keys <- tibble::lst(airlines, airports, flights, planes, weather) %>%
  as_dm()
```

# Defining Keys

Though you now have a new `dm` object, containing all your data, a key detail is missing still: 
How your five tables are connected the foreign keys) and which column(s) uniquely identify observation (the primary keys).

## Primary Keys

`dm` offers `cdm_enum_pk_candidates()` to identify primary keys and `cdm_add_pk()` to add them. 

```{r}
cdm_enum_pk_candidates(
  dm = flights_dm_no_keys,
  table = planes
)
```

Now, add any primary keys you may have identified:

```{r}
flights_dm_only_pks <- flights_dm_no_keys %>%
  cdm_add_pk(table = airlines, column = carrier) %>%
  cdm_add_pk(airports, faa) %>%
  cdm_add_pk(planes, tailnum)
flights_dm_only_pks
```

To get an overview of the primary keys after setting them, call `cdm_get_all_pks()`.

```{r}
cdm_get_all_pks(flights_dm_only_pks)
```

## Foreign Keys

```{r}
cdm_enum_fk_candidates(
  dm = flights_dm_only_pks,
  table = flights,
  ref_table = airlines
)
```

You define how your tables relate using `cdm_add_fk()`. 
First, define the tables you wish to connect to another using the function's options `table` and `ref_table`. 
Then indicate in `column` which column of `table` refers to `ref_table`'s primary key, which you've defined above. 
Use `check = FALSE` to omit consistency checks.

```{r}
flights_dm_all_keys <- flights_dm_only_pks %>%
  cdm_add_fk(table = flights, column = tailnum, ref_table = planes, check = FALSE) %>%
  cdm_add_fk(flights, carrier, airlines) %>%
  cdm_add_fk(flights, origin, airports)
flights_dm_all_keys
```

## Retrieving Keys

To retrieve your keys later on, use `cdm_get_all_fks()` or `cdm_get_fk()` for its singular version. 

```{r}
cdm_get_all_pks(flights_dm_all_keys)
```

Voilà, here's your `dm` object that you can work with:

```{r}
flights_dm_all_keys
```
