---
title: "Manipulating individual tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Technical: Manipulating individual tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

``````{r setup, include = FALSE}
source("setup/setup.R")
``````

This vignette deals with situations where you want to transform tables of your `dm` object and then update an existing table or add a new table to the `dm` object.
There are two approaches:

1. extract the tables relevant to the calculation, perform the necessary transformations, and (if needed) recombine the resulting table into a `dm`,
1. do all this within the `dm` object by zooming to a table and manipulating it.

Both approaches aim at maintaining the key relations whenever possible.
We will explore the first approach here. 
For the second approach, see `vignette("tech-dm-zoom")`.

## Enabling {dplyr}-workflow within a `dm`

The `dm_get_tables()` and `pull_tbl()` functions have a new experimental argument `keyed`, which defaults to `FALSE`.
If set to `TRUE`, a list of objects of class `dm_keyed_tbl` is returned instead.
Because `dm_keyed_tbl` inherits from `tbl` or `tbl_lazy`, many {dplyr} and {tidyr} verbs will work unchanged.
These objects will also attempt to track primary and foreign keys, so that they are available for joins and when recombining these tables later into a `dm` object.

When you are finished with transforming your data, you can use `dm()` or `new_dm()` to recombine the tables into a `dm` object.
The resulting tables in the `dm` will have all the primary and foreign keys available that could be tracked from the original table.
Reconstructing the `dm` object is not strictly necessary if you're primarily interested in deriving one or multiple separate tables for analysis.

If this workflow proves as useful as it seems, subsetting tables via `$`, `[[` will default to `keyed = TRUE` in a forthcoming major release of {dm}.


## Examples

So much for the theory, but how does it look and feel?
To explore this, we once more make use of our trusted {nycflights13} data.

### Use case 1: Add a new column to an existing table

Imagine you want to have a column in `flights`, specifying if a flight left before noon or after.
Just like with {dplyr}, we can tackle this with `mutate()`.
Let us do this step by step:

```{r zoom}
library(dm)
library(dplyr)

flights_dm <- dm_nycflights13(cycle = TRUE)
flights_dm

flights_keyed <-
  flights_dm %>% 
  dm_get_tables(keyed = TRUE)

# The print output for a `dm_keyed_tbl` looks very much like that from a normal
# `tibble`, with additional details about keys.
flights_keyed$flights

flights_tbl_mutate <-
  flights_keyed$flights %>%
  mutate(am_pm_dep = if_else(dep_time < 1200, "am", "pm"), .after = dep_time)

flights_tbl_mutate
```

To update the original `dm` with a new `flights` table we use `dm()`.
The bang-bang-bang (`!!!`) is a technical necessity that will become superfluous in a forthcoming release.

```{r zoom2}
updated_flights_dm <- dm(
  flights = flights_tbl_mutate,
  !!!flights_keyed[c("airlines", "airports", "planes", "weather")]
)

# The only difference in the `dm` print output is the increased number of
# columns
updated_flights_dm

# The schematic view of the data model remains unchanged
dm_draw(updated_flights_dm)
```

### Use case 2: Creation of a surrogate key

The same course of action could, for example, be employed to create a surrogate key for a table, a synthetic simple key that replaces a compound key.
We can do this for the `weather` table.

```{r}
library(tidyr)

flights_keyed$weather

# Maybe there is some hidden candidate for a primary key that we overlooked?
enum_pk_candidates(flights_keyed$weather)

# Seems we have to construct a column with unique values
# This can be done by combining column `origin` with `time_hour`, if the latter
# is converted to a single time zone first; all within the `dm`:
weather_tbl_mutate <-
  flights_keyed$weather %>%
  # first convert all times to the same time zone:
  mutate(time_hour_fmt = format(time_hour, tz = "UTC")) %>%
  # paste together as character the airport code and the time
  unite("origin_slot_id", origin, time_hour_fmt) %>%
  select(origin_slot_id, everything())

# check if we the result is as expected:
weather_tbl_mutate %>% 
  enum_pk_candidates() %>%
  filter(candidate)

# We apply the same transformation to create
# the foreign key in the flights table:
flights_tbl_mutate <-
  flights_keyed$flights %>%
  mutate(time_hour_fmt = format(time_hour, tz = "UTC")) %>%
  unite("origin_slot_id", origin, time_hour_fmt) %>%
  select(origin_slot_id, everything())

surrogate_flights_dm <-
  dm(
    weather = weather_tbl_mutate,
    flights = flights_tbl_mutate,
    !!!flights_keyed[c("airlines", "airports", "planes")]
  ) %>%
  dm_add_fk(flights, origin_slot_id, weather, origin_slot_id)

surrogate_flights_dm %>%
  dm_draw()
```

### Use case 3: Disentangle `dm`

If you look at the `dm` created by `dm_nycflights13(cycle = TRUE)`, you see that two columns of `flights` relate to the same table, `airports`.
One column stands for the departure airport and the other for the arrival airport.
This generates a cycle which leads to failures with many operations that only work on cycle-free data models, such as `dm_flatten_to_tbl()`, `dm_filter()` or `dm_wrap_tbl()`.
In such cases, it can be beneficial to "disentangle" the `dm` by duplicating the referred table.
One way to do this in the {dm}-framework is as follows:

```{r}
disentangled_flights_dm <-
  dm(
    destination = flights_keyed$airports,
    origin = flights_keyed$airports,
    !!!flights_keyed[c("flights", "airlines", "planes", "weather")]
  ) %>%
  # Key relations are also duplicated, so the wrong ones need to be removed
  dm_rm_fk(flights, dest, origin) %>%
  dm_rm_fk(flights, origin, destination)

disentangled_flights_dm %>% 
  dm_draw()
```


### Use case 4: Add summary table to `dm`

Here is an example for adding a summary of a table as a new table to a `dm`. Foreign-key relations are taken care of automatically.
This example shows an alternative approach of deconstruction and reconstruction using `pull_tbl()` and `dm()` with the original `dm` object.

```{r}
derived_flights_dm <-
  flights_dm() %>% 
  pull_tbl(flights, keyed = TRUE) %>% 
  count(origin, carrier) %>% 
  dm(flights_dm)

derived_flights_dm %>% 
  dm_draw()
```

### Use case 5: Joining tables

If you would like to join some or all of the columns of one table to another, you can make use of one of the `..._join()` methods for a `dm_keyed_tbl`.
In many cases, using keyed tables derived from a `dm` object allows omitting the `by` argument without triggering a message, because they are safely inferred from the foreign keys stored in the `dm_keyed_tbl` objects.
For the syntax, please see the example below.

```{r}
planes_for_join <-
  flights_keyed$planes %>%
  select(tailnum, plane_type = type)

joined_flights_tbl <-
  flights_keyed$flights %>%
  # let's first reduce the number of columns of flights
  select(-dep_delay:-arr_delay, -air_time:-minute, -starts_with("sched_")) %>%
  # in the {dm}-method for the joins you can specify which columns you want to
  # add to the subsetted table
  left_join(planes_for_join)

joined_flights_dm <- dm(
  flights_plane_type = joined_flights_tbl,
  !!!flights_keyed[c("airlines", "airports", "weather")]
)

# this is how the table looks now
joined_flights_dm$flights_plane_type

# also here, the FK-relations are transferred to the new table
joined_flights_dm %>% 
  dm_draw()
```
