---
title: "Filtering in relational data models"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Filtering}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
fansi::set_knit_hooks(knitr::knit_hooks)
options(crayon.enabled = TRUE, width = 75, cli.width = 75)

knit_print.grViz <- function(x, ...) {
  x %>%
    DiagrammeRsvg::export_svg() %>%
    c("`````{=html}\n", ., "\n`````\n") %>%
    knitr::asis_output()
}
```

The {dm} package offers functions to work with relational data models in R.

This document introduces you to the filtering functions and shows you how to apply them to data that is seperated into multiple tables.

The example is working with data from the {nycflights13} package, that contains tables linked with each other.

First, we load the packages we need:

```{r message=FALSE, warning=FALSE}
library(dm)
library(tidyverse)
library(nycflights13)
```

## Data: nycflights13

To explore filtering with `dm`, we'll use the {nycflights13} data with its tables `flights`, `planes`, `airlines` and `airports`.

This dataset contains all 336 776 flights that departed from New York City in 2013 with 3322 different planes and 1458 airports involved.
The data comes from the US Bureau of Transportation Statistics, and is documented in `?nycflights13::flights`.

First of all, we have to create a `dm` object from the {nycflights13} data.
The built-in `dm::cdm_nycflights13()` function takes care of this.

A [data model object](https://krlmlr.github.io/dm/articles/dm-class-and-basic-operations.html#class-dm) contains the data as well as metadata.

If you want to create a `dm` from other tables, please use the `new_dm()` and `dm()` functions.

```{r}
dm <- cdm_nycflights13()
```

A `dm` object output with data and metadata consists of three elements: 

1. the table sources,
2. the data model,
3. and a review of active filter conditions.

The console output is colored for a clear output:

```{r}
dm
```

Now we know that there are five tables in our `dm` object.
But how are they connected?
These relations are best displayed as a visualization of the entity-relationship model:

```{r}
cdm_draw(dm)
```

You can look at a single table with `tbl`.
Print the `airports` table:

```{r}
tbl(dm, "airports")
```

## Filtering a `dm` object {#filter}

`cdm_filter()` allows you to select a subset of a `dm` object.

### How it works

Filtering a `dm` object is not that different from filtering a dataframe or tibble with `dplyr::filter()`.

The corresponding `dm` function is `dm::cdm_filter()` and creates a subset of one table depending on the input.
But `dm` does not stop here: A cascade of joins is performed along the foreign key relations.
This results in a new `dm` containing only rows related to the filter operation.

Currently, this only works if the graph induced by the foreign key relations is cycle free.
Luckily, this is the default for `cdm_nycflights13()`.

### Filtering Examples

Let's see it in action:

**We only want the data that is related with John F. Kennedy International Airport.**

The total number of rows drops from  367 687 to 138 786 (apart from the disconnected `weather` table).

You can see this in the `Rows`-part of the print output.

```{r}
dm %>% 
  cdm_filter(airports, name == "John F Kennedy Intl")
```

Another way to check the number of rows is the function `cdm_nrow()`:

```{r}
dm %>%
  cdm_filter(airports, name == "John F Kennedy Intl") %>%
  cdm_nrow()
```

Next example:

**Get a `dm` object containing data for flights from New York to the Dulles International Airport in Washington D.C., abbreviated with `IAD`.**

```{r}
dm %>% 
  cdm_filter(flights, dest == "IAD")
```

Chaining multiple filters on different tables is also supported.

An Example:

**Get all flights from American Airlines which didn't depart from John F. Kennedy International Airport in May 2013.**

```{r}
dm_aa_may <- dm %>%
  cdm_filter(airlines, name == "Delta Air Lines Inc.") %>%
  cdm_filter(airports, name != "John F Kennedy Intl") %>%
  cdm_filter(flights, month == 5)
```

You can inspect the filtered tables with `tbl`.

In the `airlines` table, Delta is the only remaining carrier:

```{r}
dm_aa_may %>% 
  tbl("airlines")
```

Which planes were used to service these flights?

```{r}
dm_aa_may %>% 
  tbl("planes")
```

And indeed, there are only flights from `month == 5`:

```{r}
dm_aa_may %>% 
  tbl("flights")
```

For comparison, let's review the equivalent manual query for the `flights` in `dplyr` syntax:

```{r}
flights %>%
  left_join(airlines, by = "carrier") %>%
  filter(name == "American Airlines Inc.") %>%
  left_join(airports, by = c("origin" = "faa")) %>%
  filter(
    name.y != "John F Kennedy Intl",
    month == 5
  )
```

The `dm` code is leaner, because the foreign key information is encoded in the object.

## SQL statements behind filtering a `dm` object on a database

{dm} is meant to work with relational databases, locally as well on databases.
In your project the data is probably not stored locally, but in a remote [relational database](https://krlmlr.github.io/dm/articles/dm-introduction-relational-data-models.html#relational-databases) which can be queried with SQL statements.

You can check the queries by using the `sql_render()` from the [`dbplyr` package](https://dbplyr.tidyverse.org/).

Example:

**Print the SQL statements for getting all flights from American Airlines which didn't depart from John F. Kennedy International Airport in May 2013 when the data is stored in a sqlite database.**

To show the SQL query behind a `cdm_filter()`, we copy the `flights`, `airlines` and `airports` tables from the `nyflights13` dataset to a temporary in-memory database using the built-in function `cdm_copy_to()` and `dbplyr::src_memdb`.

Then we filter the data and print the corresponding SQL statement with `dbplyr::sql_render()`.

```{r, warning=F}
dm %>%
  cdm_select_tbl(flights, airlines, airports) %>%
  cdm_filter(flights, month == 5) %>% 
  cdm_copy_to(dbplyr::src_memdb(), ., unique_table_names = TRUE) %>% 
  cdm_filter(airlines, name == "American Airlines Inc.") %>%
  cdm_filter(airports, name != "John F Kennedy Intl") %>%
  cdm_apply_filters() %>% 
  cdm_get_tables() %>%
  map(dbplyr::sql_render)
```

Further reading: {dm}'s function for copying data [from and to databases](https://krlmlr.github.io/dm/articles/dm.html#copy).
