---
title: "Introduction to dm"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dm overview}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
fansi::set_knit_hooks(knitr::knit_hooks)
options(crayon.enabled = TRUE, width = 75, cli.width = 75)

knit_print.grViz <- function(x, ...) {
  x %>% 
    DiagrammeRsvg::export_svg() %>% 
    c("`````{=html}\n", ., "\n`````\n") %>% 
    knitr::asis_output()
}
```

The goal of both the package dm and the class `dm` - which comes along with it - is to make your life easier when you have to deal with data models the data therein.

Let's first have a look at the new class:

## Class `dm`

The `dm` class contains a set of tables as well as information about their primary keys and the relations between the tables. You can have your `dm` object on different data sources, such as a local environment or a database. In the latter use case you need to have the package dbplyr installed. 

`dm` is a wrapper around two classes, which are independent from each other:

1. part of the object is a [`src`-object](https://dplyr.tidyverse.org/reference/src.html) from dplyr
2. the other part is a class `data_model` object
 from the package [datamodelr](https://github.com/bergant/datamodelr)^[Use: `devtools::install_github("https://github.com/bergant/datamodelr")` to install it]

The `src`-object contains the information, where the tables of your data model live. And in the `data_model` object the meta-information is stored:

- what are the primary keys in your schema?
- how are the tables related to each other, i.e. what are the foreign key relations?

Last but not least, a third part of the object is a named list containing the tables on the `src`.

So much to the theory, let's see how to construct such an object and how it looks in `R`:

## Examples of `dm` objects {#ex_dm}

Relevant functions for creating `dm` objects are:

1. `dm()`
2. `as_dm()`

We can use the tables from the famous package {nycflights13}:
```{r message=FALSE}
library(dm)
library(dplyr)
flights_dm <- dm(src_df(pkg = "nycflights13"))
flights_dm
```

Here we make use of the fact, that the function `dm(src, data_model = NULL)` includes all available tables on a source in the `dm` object, should the `data_model`-argument be left `NULL`. 
Thus you can use this for example on a postgres database that you access via `src_postgres()` (with the appropriate arguments `dbname`, `host`, `port`, ...), to produce a `dm` object with all the tables on the database.

Another way of creating a `dm` object is calling `as_dm()` on a list of `tbls`:
```{r}
iris_dm <- as_dm(list("iris1" = iris, "iris2" = iris))
iris_dm
```

## Methods for `dm` objects

tbl(dm, "table_name") etc. (copy_to...) FIXME: complete

But how can we use dm-functions to manage primary keys of the tables in a `dm` object?

## Primary keys of `dm` objects {#pk}

Useful functions for managing primary key settings are:

1. `cdm_add_pk()`
2. `cdm_has_pk()`
3. `cdm_get_pk()`
4. `cdm_rm_pk()`
5. `cdm_enum_pk_candidates()`

As of yet `dm` objects only support one-column primary keys. If your tables have unique compound keys, maybe consider adding a surrogate key column. 
If you created a `dm` object according to the examples in ["Examples of `dm` objects"](#ex_dm), your object does not yet have any primary keys set.
So let's add one.
By default `cdm_add_pk()` checks if the column of the table given by the user is a unique key.
Since the `iris` dataset does not have any unique one-column keys, we will use the `nycflights13` tables, i.e. `flights_dm`.

```{r}
cdm_has_pk(flights_dm, airports)
flights_dm_with_key <- cdm_add_pk(flights_dm, airports, faa)
flights_dm_with_key
```
Still looks exactly the same as before... let's check:
```{r}
cdm_has_pk(flights_dm_with_key, airports)
```
Get the name of the column that is marked as primary key of the table:
```{r}
cdm_get_pk(flights_dm_with_key, airports)
```
Remove a primary key by:
```{r}
cdm_rm_pk(flights_dm_with_key, airports) %>% 
  cdm_has_pk(airports)
```

If you still need to get to know your data better, and it is already available in the form of a `dm` object, you can use the function `cdm_enum_pk_candidates()` in order to get information, which columns of a table are unique keys:
```{r}
cdm_enum_pk_candidates(flights_dm_with_key, airports)
```
Table `flights` does not have any one-column primary key candidates:
```{r}
cdm_enum_pk_candidates(flights_dm_with_key, flights) %>% count(candidate)
```

## Foreign keys

Useful functions for managing foreign key relations are:

1. `cdm_add_fk()`
2. `cdm_has_fk()`
3. `cdm_get_fk()`
4. `cdm_rm_fk()`
5. `cdm_enum_fk_candidates()`

Now it gets (even more) interesting: we want to define relations between different tables.
With the function `cdm_add_fk()` you can define, which column of which table points to another table's column.

When dealing with data models, a foreign key from one table has to always point to a primary key of another table.
The primary key of the referred table must be set with `cdm_add_pk()`.
`cdm_add_fk()` will find the primary key column of the referenced table by itself and make the indicated column of the child table point to it.

```{r}
flights_dm_with_key %>% cdm_add_fk(flights, origin, airports)
```
This will throw an error:
```{r error=TRUE}
flights_dm %>% cdm_add_fk(flights, origin, airports)
```
Let's create a `dm` object with a foreign key relation to work with later on:
```{r}
flights_dm_with_fk <- cdm_add_fk(flights_dm_with_key, flights, origin, airports)
```
What if we try to add another foreign key relation from `flights` to `airports` to the object? Column `dest` might work, since it also contains airport codes:
```{r error=TRUE}
flights_dm_with_fk %>% cdm_add_fk(flights, dest, airports)
```
As you can see, behind the scenes, checks are executed automatically by the functions of `dm` to prevent steps that do not comply with the logic of data models.

Use `cdm_has_fk()` for checking if a foreign key exists, which is pointing from one table to another:
```{r}
flights_dm_with_fk %>% cdm_has_fk(flights, planes)
flights_dm_with_fk %>% cdm_has_fk(flights, airports)
```

If you want to access the name of the column, which acts as a foreign key of one table to another table's column, use `cdm_get_fk()`:
```{r}
flights_dm_with_fk %>% cdm_get_fk(flights, planes)
flights_dm_with_fk %>% cdm_get_fk(flights, airports)
```

Remove foreign key relations with `cdm_rm_fk()` (parameter `column = NULL` means removal of all relations):
```{r error=TRUE}
flights_dm_with_fk %>% 
  cdm_rm_fk(table = flights, column = dest, ref_table = airports) %>% 
  cdm_get_fk(flights, airports)
flights_dm_with_fk %>% 
  cdm_rm_fk(flights, origin, airports) %>% 
  cdm_get_fk(flights, airports)
flights_dm_with_fk %>% 
  cdm_rm_fk(flights, NULL, airports) %>% 
  cdm_get_fk(flights, airports)
```

Another function for getting to know your data better (cf. `cdm_enum_pk_candidates()` in ["Primary keys of `dm` objects"](#pk)) is `cdm_enum_fk_candidates()`. Use it to get an overview over foreign key candidates pointing from one table to another:
```{r}
cdm_enum_fk_candidates(flights_dm_with_key, weather, airports)
```

## Visualising the data model

Relevant functions for visualising the data model:^[The functions in this section rely heavily on related functions of {datamodelr}.]

1. `cdm_set_colors()`
1. `cdm_get_colors()`
1. `cdm_get_available_colors()`
1. `cdm_draw()`

Once you have all primary keys set and all foreign key relations defined, it might be interesting to look at a graphical representation of your data model.

Let's first create a slightly more interconnected `dm` object from the {nycflights13} package tables:
```{r}
flights_dm_w_many_keys <- 
  flights_dm %>% 
  cdm_add_pk(planes, tailnum) %>% 
  cdm_add_pk(airports, faa) %>% 
  cdm_add_pk(airlines, carrier) %>% 
  cdm_add_fk(flights, origin, airports) %>% 
  cdm_add_fk(flights, carrier, airlines) %>% 
  cdm_add_fk(flights, tailnum, planes, check = FALSE) %>% 
  cdm_add_fk(flights, dest, airports, check = FALSE)
flights_dm_w_many_keys
```
You probably realized the option `check = FALSE` in the last two function calls when adding foreign keys.
The check normally ensures, that the set of values of the column of the first table is a subset of the set of values of the primary key column of the referenced table.
Since in our case this isn't the case, but we want complexity, we turned the check off.
It is recommended not to do this, since you might run into errors or other unwanted effects, mainly when working on databases.

You can use colors to visually group your tables into families (which could reflect their logical grouping).
Currently, there are only eight different colors available (plus the option of having borders displayed around the tables or not).
To find out which colors exist, use `cdm_get_available_colors()`:
```{r}
cdm_get_available_colors()
```

The assignment is carried out with `cdm_set_colors()`, using a syntax similar to `switch()`.
Any table you don't specifically assign will be displayed in the default color.

Use `cdm_set_colors()` to assign the respective colors. 
The result of `cdm_set_colors()` is a `dm` object.
The information about the color is stored in its `data_model`-part.
```{r}
flights_dm_w_many_keys_and_colors <-
  flights_dm_w_many_keys %>% 
  cdm_set_colors(
    flights = "blue",
    airlines = ,
    airports = "orange",
    planes = "green_nb"
  )
```

Draw the schema with `cdm_draw()`. 
```{r}
cdm_draw(flights_dm_w_many_keys_and_colors)
```

The colors can be queried with `cdm_get_colors()`. 
```{r}
cdm_get_colors(flights_dm_w_many_keys_and_colors)
```

There are quite a few options for customization available. 
Please read the documentation to find out more.
One argument of the function we'd like to draw your attention to, though, is `table_names`.
This argument accepts a character vector with the table names which are to be printed, in case you would like to only see a subset of the whole data model.
```{r}
cdm_draw(flights_dm_w_many_keys_and_colors, table_names = c("flights", "airports", "planes"))
```

## Joining tables

## Filtering a `dm` object
